---
title: "Portable derivative provenance and methods text"
author: "Michael Hallquist"
date: "11 Sep 2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Portable derivative provenance and methods text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Native `postprocess_subject()` and `extract_rois()` outputs receive portable
records automatically. A NIfTI or TSV derivative has an adjacent JSON sidecar
and a companion directory with the same stem and a `.provenance` suffix:

```text
sub-01_task-rest_desc-clean_bold.nii.gz
sub-01_task-rest_desc-clean_bold.json
sub-01_task-rest_desc-clean_bold.provenance/
  methods.md
  bibliography.bib
  attempt-<record-id>.json
  upstream/...
```

The JSON describes the derivative's producer. Each attempt has a separate
record, so requesting different settings while reusing a file cannot rewrite
its generating history. ROI time series, connectivity matrices, and optional
voxel-retention diagnostics each have their own sidecar and companions.
`extract_rois()` returns companion paths in its `provenance` element; extraction
job output manifests include those paths as well as the TSV outputs.

## Inspect and share a derivative

```{r inspect, eval=FALSE}
library(BrainGnomes)

# Use the file returned by postprocess_subject(), or an ROI TSV from extract_rois().
derivative <- "/study/postproc/sub-01/func/sub-01_task-rest_desc-clean_bold.nii.gz"
metadata <- read_derivative_provenance(derivative)
record <- metadata$BrainGnomes
record$Resolved
record$Operations
record$Execution
cat(record$MethodsText)

# The destination must be new. The export validates the derivative and saved
# companions, copies them, and adds a derivative dataset description.
bundle <- export_derivative_provenance(derivative, "/shared/sub-01-clean-export")
read_derivative_provenance(bundle$derivative)
```

Copy the whole exported directory to a collaborator. Reading its JSON requires
neither SQLite nor the original study directory. The record includes original
source paths, MD5 checksums, sizes, selected upstream metadata, and available
execution identities; these remain interpretable when source paths are offline.
Original images, container images, and atlas voxel arrays are identified rather
than copied. Source paths and subject identifiers are retained, so the bundle
is not an anonymization tool.

`read_derivative_provenance()` verifies the derivative checksum and schema. It
can read a sidecar even when its companion directory is unavailable.
`export_derivative_provenance()` additionally checks the presence, confinement,
and checksums of saved report/citation assets and generated prose. An incomplete
or modified companion directory is an error. It publishes the new directory only
after these checks succeed and refuses to overwrite an existing export.
Only declared companions and derivative-attempt JSON files are included;
unrelated additions to the companion directory are excluded.

Optional external identifiers can supplement the portable record:

```{r external-identifiers, eval=FALSE}
export_derivative_provenance(
  derivative, "/shared/sub-01-clean-with-datalad",
  datalad = list(dataset_id = "<existing-dataset-id>",
                 commit = "<existing-commit>", annex_key = "<existing-annex-key>")
)
```

These identifiers are explicitly marked `user_supplied_not_verified`.
BrainGnomes neither invokes DataLad nor creates or verifies a dataset or commit.

## Version 1 record contract

`BrainGnomes.SchemaVersion` is `brain-gnomes-derivative-v1`. The extension is
BrainGnomes metadata, not a claim to implement a standardized processing-history
schema. Readers reject unsupported schema versions. Arrays remain arrays when
empty or when they contain a single element. Missing optional records are null
or explicitly marked unavailable, rather than inferred from a filename.

| Field within `BrainGnomes` | Meaning |
| --- | --- |
| `RecordID`, `RecordedAt`, `Stage`, `Status` | Unique derivative-attempt identity, UTC recording time, native stage, and overall outcome. |
| `Derivative` | Output identity and checksum; paths describe the original publication location. |
| `Requested`, `RequestedSequence` | Selected submitted configuration and enabled sequence before execution-time omissions. |
| `Resolved.ProcessingSequence`, `Operations` | Resolved image sequence and ordered operation states, parameters, and evidence. Repeated operations have separate entries. |
| `Resolved.Confounds` | Expanded processed/unprocessed selections, constant-column exclusions, written regression design columns in order, calculated columns, motion-filter settings, and transformations actually applied to processed confounds. |
| `Resolved.CensorPolicy`, `Resolved.Censor` | Requested censor rule and resolved retained/excluded indices. Indices are one-based; 1 retains a volume and 0 censors it. ROI records distinguish applying censoring from recognizing an already-censored input. |
| `Resolved.TR`, `Resolved.Space`, `InputGeometry`, `OutputGeometry` | Repetition time, BIDS space label, and NIfTI dimensions, spacing, units, and transform fields when available. |
| `Resolved.IntensityConvention`, `Resolved.IntensityReference` | Input-scale convention and normalization calibration when normalization ran. ROI records describe the inherited scale and PCA convention. |
| `Resolved.Atlas`, `Resolved.RegionDiagnostics` | Original and resolved atlas identities, label values, grid and resampling details, and regional retention decisions. The atlas file itself is not archived. |
| `SourceFiles`, `SourceMetadata`, `SourceDataset` | Source identities and supplied metadata, including a parent BrainGnomes record when present. |
| `Software`, `Execution` | BrainGnomes/R/loaded-package identities, configured container identities, and available run, work-unit attempt, manifest, receipt, and scheduler-task records. |
| `AttemptFile`, `ReusedRecordID` | Relative attempt link for a producer and producer identity for a reuse attempt, respectively. |
| `Validation`, `Error` | Available validation results and attempt failures. Completion and validation are separate outcomes. |
| `MethodsText`, `BibliographyParts`, `HumanReviewRequired` | Draft methods, bibliography blocks, and an always-true human-review requirement. |
| `Assets`, `Companions`, `ExportedAttempts`, `ExternalReferences` | Relative saved-asset links and checksums, generated-text identities, attempts included in an export, and optional external references. |

A job's `Execution.Receipt` is the startup runtime-verification receipt defined
by the job contract. It is not proof of successful processing. Array jobs may
share a contract/receipt; the current scheduler task is recorded separately.
Calls outside a scheduled run still produce derivative provenance, with missing
run/contract/receipt fields left null. Available records are embedded so their
meaning does not depend on reopening the database or original logs.

The sidecar preserves applicable `SpatialReference` and postprocessing
`RepetitionTime` metadata. Exports add `dataset_description.json` with
`DatasetType` and `GeneratedBy`. When original dataset roots can be identified,
`Sources` contains BIDS URIs whose keys resolve through `DatasetLinks`; original
absolute paths remain in the extension regardless of source accessibility.
These fields follow the [BIDS derivative metadata guidance](https://bids-specification.readthedocs.io/en/v1.11.1/derivatives/common-data-types.html)
and [dataset description specification](https://bids-specification.readthedocs.io/en/v1.11.1/modality-agnostic-files/dataset-description.html).
The bundle is a transport layout for an individual derivative; these fields do
not establish that its layout or every ROI suffix forms a complete,
validator-compliant BIDS dataset.

## Interpret execution and reuse

- A `requested` record contains an intended configuration. `running` is a
  transient operation state; it becomes `executed` after the operation returns.
- A `completed` producer describes a written derivative. Validation results may
  still indicate failure when the configured policy allows processing to continue.
- A `failed` attempt records its error and marks in-flight operations failed.
  Previously executed operations retain their state. An old sidecar whose image
  was removed or changed during replacement cannot pass checksum validation.
- A `skipped` operation or attempt is not described as successful processing.
  Examples include unavailable mask/regressor selection and connectivity with
  fewer than 20 retained timepoints.
- Reusing a final derivative retains its verified producer sidecar and writes a
  new `reused` attempt. Reused intermediates carry their file identity and mark
  current parameters `requested_only`; they do not establish the intermediate's
  original settings. Inspect those origins before accepting generated prose.
- An existing derivative with no BrainGnomes producer record receives `unknown`
  provenance, with generating settings and original sources left unknown. The
  current request is retained only in the separate reuse attempt.

## Review generated methods and citations

The paragraph is generated from resolved records and executed operations, with
explicit qualifications for reuse and unsuccessful validation. Review it against
the sidecar and study protocol before using it in a publication. Add scientific
rationale, acquisition details, quality-control decisions, atlas citations, and
any externally performed processing that these records cannot establish.

Available upstream `CITATION.bib`, `CITATION.md`, `CITATION.html`, and
`CITATION.tex` files are retained verbatim, along with the participant HTML
report and locally referenced resources. Externally hosted report assets are
not downloaded. Parent BrainGnomes methods and bibliography are retained through
ROI extraction. Missing upstream citations/reports cannot be reconstructed; check
that the expected assets are present before sharing.

The bibliography includes the BrainGnomes package citation and relevant native
operation citations, such as [FSL](https://doi.org/10.1016/j.neuroimage.2011.09.015)
and [ICA-AROMA](https://doi.org/10.1016/j.neuroimage.2015.02.064), plus supplied
upstream BibTeX. When dataset metadata identifies fMRIPrep and its core paper is
absent from supplied BibTeX, the exporter adds the
[fMRIPrep methods paper](https://doi.org/10.1038/s41592-018-0235-4).
Upstream bibliography blocks are preserved rather than rewritten, so human
review should also resolve duplicate citation keys or entries and add any
missing tool-, atlas-, or version-specific references.
