---
title: "Diagnosing Pipeline Runs"
author: "Michael Hallquist & Dan Shallal"
date: "30 Aug 2026"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Diagnosing Pipeline Runs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

Running an fMRI pipeline with BrainGnomes typically means launching many jobs:
BIDS conversion, MRIQC, fMRIPrep, ICA-AROMA, one or more postprocessing streams,
and optional ROI extraction. Some jobs must wait for earlier jobs to finish.
Each job writes logs and completion markers, and a project can have many runs.
Without a run-specific view, it is hard to answer even basic questions:

- What already succeeded?
- What is still running or queued?
- What failed, and was it the *root cause* or just downstream of another
  failure?
- Where should I look next in the logs?

BrainGnomes provides two useful levels of information:

- `get_project_status()` gives a project-wide table of step-level completion
  flags and times for all subjects.
- `get_subject_status()` focuses that same information on a single subject
  (and session), making it easy to see where one subject is stuck.
- `get_project_runs()`, `get_run_jobs()`, `diagnose_project()`, and
  `find_run_logs()` inspect one submitted run without prompts.
- `diagnose_pipeline()` provides the established interactive browser when you
  want to follow job relationships and read logs step by step.

This vignette walks through a practical workflow that starts with the
high-level status summaries, narrows the problem to one run, and then shows how
to inspect provenance, retry failed work, or cancel work that is still active.

## Load the project

```{r, eval=FALSE}
library(BrainGnomes)

project_dir <- "/project/my_study"
scfg <- load_project(project_dir)
```

Loading a project validates its configuration without opening the setup wizard
or changing the saved YAML file. This is the same configuration object used by
`run_project()` and the diagnosis functions below.

# Choose the run you mean

Every submitted `run_project()` call returns a run handle. Its `run_id` is a
stable label that connects the jobs, logs, diagnosis, and provenance from that
submission.

```{r, eval=FALSE}
run <- run_project(scfg)
run_id <- run$run_id
```

In a later R session, list previous runs and copy the ID you want to inspect:

```{r, eval=FALSE}
runs <- get_project_runs(scfg)
runs
run_id <- runs$run_id[[1]]
```

Most run-based functions also accept `run_id = "latest"`. That shortcut is
convenient for inspection. For retry or cancellation, an explicit ID is safer
because another submission could otherwise become the latest run.

# Project-level triage with `get_project_status()`

Start with a table of completion flags and times for all subjects:

```{r, eval=FALSE}
status_df <- get_project_status(scfg)
status_df
summary(status_df)
```

Captured output (16 Feb 2026):

```
rows: 10 cols: 12

sub_id  bids_conversion_complete  mriqc_complete  fmriprep_complete  aroma_complete  task_complete
540294  TRUE                     TRUE            FALSE              FALSE           FALSE
540295  TRUE                     TRUE            FALSE              FALSE           FALSE
...     ...                      ...             ...                ...             ...
540311  TRUE                     TRUE            FALSE              FALSE           FALSE

Summary:
step                       n_complete
bids_conversion_complete   10
mriqc_complete             10
fmriprep_complete           0
aroma_complete              0
task_complete               0
```

This gives a quick pass/fail style view. In this run, BIDS conversion and MRIQC
are complete for all subjects, while downstream steps are not yet complete.

Under the hood, `get_project_status()` and `get_subject_status()`:

- look in the project’s BIDS directory to discover which subjects/sessions
  exist, and
- scan each subject’s log directory for BrainGnomes `.complete` (and related)
  marker files.

From these filesystem markers, they derive strict `*_complete` flags and
completion times for each enabled step.

You can also scan log markers for explicit failures:

```{r, eval=FALSE}
list.files(
  scfg$metadata$log_directory,
  pattern = "_fail$",
  recursive = TRUE,
  full.names = FALSE
)
```

Captured output:

```
[1] "sub-540295/.fmriprep_sub-540295_fail"
[2] "sub-540300/.fmriprep_sub-540300_fail"
[3] "sub-540301/.fmriprep_sub-540301_fail"
[4] "sub-540302/.fmriprep_sub-540302_fail"
[5] "sub-540303/.aroma_sub-540303_fail"
```

# Subject-level check with `get_subject_status()`

To check one subject:

```{r, eval=FALSE}
get_subject_status(scfg, "540296")
```

Captured output:

```
  sub_id ses_id bids_conversion_complete bids_conversion_time mriqc_complete
1 540296   <NA>                     TRUE  2026-02-14 18:27:01           TRUE
           mriqc_time fmriprep_complete fmriprep_time aroma_complete aroma_time
1 2026-02-15 04:00:02             FALSE          <NA>          FALSE       <NA>
  task_complete task_time
1         FALSE      <NA>
```

This is often the fastest way to answer, "Where is subject X stuck?"

# Inspect one run without prompts

Use `get_run_jobs()` to see every tracked job in the selected run. The returned
table includes the processing step, subject, scheduler job number, timing, and
current status.

```{r, eval=FALSE}
jobs <- get_run_jobs(scfg, run_id)
jobs[, c("job_id", "job_name", "status", "time_submitted", "time_ended")]
```

For a shorter failure-focused report, use `diagnose_project()`. It separates
jobs that failed, were cancelled, or could not run because an earlier job
failed, and it locates their output and error logs when those files are present.

```{r, eval=FALSE}
diagnosis <- diagnose_project(scfg, run_id)
diagnosis$failures
diagnosis$logs

# Or ask only for the log-file locations.
failed_logs <- find_run_logs(scfg, run_id, failed_only = TRUE)
```

Before deciding what to rerun, inspect the settings that produced the run:

```{r, eval=FALSE}
provenance <- get_run_provenance(scfg, run_id)
provenance$request
provenance$execution$subjects
provenance$configuration$snapshot_file
```

The provenance record contains the resolved stages and subjects, a copy of the
project configuration, requested computing resources, software and host
details, and checksums that identify containers and other files that controlled
the run. This makes it easier to answer, "What was different about this run?"

# Interactive diagnosis with `diagnose_pipeline()`

Running a full fMRI preprocessing + postprocessing pipeline generates many jobs,
log files, and potential failure points. Manually chasing SLURM job IDs and log
files quickly becomes unmanageable. The goal of `diagnose_pipeline()` is to
wrap all of this job-tracking information into a single, interactive view.

Behind the scenes, every job submitted by `run_project()` is recorded in the
project's job-tracking database. Each record stores:

- the **run ID** (called a sequence ID in some older prompts and records),
- the **subject**, job name, scheduler ID, and parent/child relationships,
- timestamps (submitted, started, finished),
- scheduler options and exit codes,
- and a derived **status** (`COMPLETED`, `STARTED`, `QUEUED`, `FAILED`,
  `FAILED_BY_EXT`).

`diagnose_pipeline()` reads this job-tracking database, reconstructs the job
dependency tree, and then guides you through it interactively. The function
returns that raw tree structure invisibly so you can also inspect it yourself:

```{r, eval=FALSE}
tree <- diagnose_pipeline(scfg)
print(tree)
```

Each node in `tree` corresponds to a single tracked job and includes its
subject, status, scheduler metadata, and parent/child links. This is the same
structure BrainGnomes uses internally to understand job dependencies.

## Run-level diagnosis

The first prompt asks whether to diagnose by **subject** or by **sequence ID**.
Here, sequence ID means the same thing as run ID:

```
Enter 1 for subject summary or 2 for sequence ID
> 2

Enter which pipeline run to diagnose. The default is the most recent.
> [Enter]
```

Choosing the run-first path lets you pick a specific pipeline run (for
example, if you re‑ran the project with a different configuration). Once you
select a run, `diagnose_pipeline()` shows all **top-level jobs** for that
run, grouped by subject, with their current status:

```
Postprocess stream1
  postprocess_stream1_sub-540294                  [STARTED]
  ---postprocess_sub-540294_task-ridl_run-01_...   [FAILED]
```

For postprocessing, this view expands nested structure: the subject‑level
postprocess job has **child jobs** for each postprocess stream, and those in
turn have children for each image‑level job (e.g., individual BOLD runs).
At a glance you can see which specific image(s) caused a stream or subject to
fail.

From the run-level view you can choose a job and drill down:

- show a detailed summary (times, parent/child jobs, exit codes),
- open the associated stdout or stderr log in the console,
- or return the log contents as a character vector for further inspection.

This is usually the fastest way to answer, *“Why did this particular run of the
pipeline fail?”* and *“Which exact job needs debugging?”*

## Subject-level diagnosis across runs

If you already know which subject you care about, you can start with the
subject‑summary branch:

```
Enter 1 for subject summary or 2 for sequence ID
> 1

Found 10 subject(s) across all runs.
Enter the number corresponding to the subject you want to view
> 3
```

Here, `diagnose_pipeline()` looks across **all runs** in the job-tracking
database and shows, for the chosen subject, which combination of runs produced
the best overall completion pattern. This helps with questions like:

- Did subject `sub-540303` ever complete fMRIPrep + AROMA + postprocessing?
- If so, which run and configuration produced that success?

The summary groups jobs by step (BIDS conversion, MRIQC, fMRIPrep, AROMA,
postprocessing streams, etc.) and highlights, for each step, which run
had the most successful outcome. You can then jump from this high‑level
summary back into run- or job-level inspection for detailed debugging.

Example subject‑level summary for `sub-540303`:

```
BIDS Conversion
  ✓ bids_conversion_sub-540303 (job 15135998, sequence 18303423-c037-477a-91af-b6289525dad2) [COMPLETED]

fMRIPrep
  • fmriprep_sub-540303 (job 15136000, sequence 18303423-c037-477a-91af-b6289525dad2) [STARTED]
```

## Interpreting job statuses

`diagnose_pipeline()` uses the following job‑level statuses:

- `COMPLETED`: the job finished successfully (based on scheduler status and
  BrainGnomes tracking).
- `STARTED`: the job has begun running or previously ran without a terminal
  status yet recorded.
- `QUEUED`: the scheduler knows about the job, but it is waiting on
  dependencies or resources.
- `FAILED`: the job itself failed (non‑zero exit code or explicit failure
  marker).
- `FAILED_BY_EXT`: the job could not run because an earlier required job
  failed. For example, postprocessing cannot run if its required fMRIPrep job
  failed.

This distinction is important: start with a `FAILED` job, because that is often
the cause. A `FAILED_BY_EXT` job is usually affected downstream work rather
than the original problem.

In contrast, `get_project_status()` and `get_subject_status()` report simple
`*_complete` flags per step; these are stricter than raw scheduler status and
rely on BrainGnomes completion markers and expected outputs. A common workflow
is:

1. Use `get_project_status()` or `get_subject_status()` for a quick overview of
   which steps are done. These functions do **not** consult the job-tracking
   database; instead they scan the project’s BIDS and log directories for `.complete`
   marker files and expected outputs, then derive strict `*_complete` flags and
   completion times per step.
2. Use `diagnose_project()` and `find_run_logs()` for a prompt-free report of one
   run, or `diagnose_pipeline()` for guided interactive investigation. These
   functions read the job-tracking database to connect jobs, statuses, and logs.

# Retry failed work after correcting the cause

A retry creates a **new run**. It does not restart scheduler jobs in place and
does not change the original run. By default it includes stages and subjects
whose jobs are `FAILED` or `CANCELLED`. BrainGnomes tells the new run to rerun
that selected work even if an old completion marker would normally skip it.

Always preview the retry first:

```{r, eval=FALSE}
retry_plan <- retry_project_run(scfg, run_id, dry_run = TRUE)
retry_plan$jobs
retry_plan$subjects
```

The preview is a plan only; it submits no jobs. Check that its stages, subjects,
and streams match what you intend. After correcting the underlying problem,
submit the retry explicitly:

```{r, eval=FALSE}
retry_run <- retry_project_run(scfg, run_id, dry_run = FALSE)
retry_run$run_id
get_run_provenance(scfg, retry_run$run_id)$invocation$parent_run_id
```

Jobs marked `FAILED_BY_EXT` are excluded by default because they were blocked
by an earlier failure rather than failing themselves. Use
`include_blocked = TRUE` when you want the same new run to include those
affected downstream stages as well:

```{r, eval=FALSE}
retry_plan <- retry_project_run(
  scfg, run_id, include_blocked = TRUE, dry_run = TRUE
)
```

# Cancel work that is still active

Cancellation applies only to tracked jobs that are queued or running. It does
not delete data, logs, or completed outputs. Preview the scheduler commands
first, then make a separate explicit call if they are correct:

```{r, eval=FALSE}
cancel_project_run(scfg, run_id, dry_run = TRUE)
cancel_project_run(scfg, run_id, dry_run = FALSE)
```

# The same recovery workflow from the command line

The command line requires either `--dry-run` for a preview or `--yes` for an
action. Use an explicit run ID for retry and cancellation:

```bash
BrainGnomes status /project/my_study --runs
BrainGnomes diagnose /project/my_study --run=<run-id>
BrainGnomes logs /project/my_study --run=<run-id> --failed-only --tail=50
BrainGnomes provenance /project/my_study --run=<run-id> --format=json

BrainGnomes retry /project/my_study --run=<run-id> --dry-run
BrainGnomes retry /project/my_study --run=<run-id> --yes

BrainGnomes cancel /project/my_study --run=<run-id> --dry-run
BrainGnomes cancel /project/my_study --run=<run-id> --yes
```
