---
title: "Inspecting and Diagnosing Pipeline Runs"
author: "Michael Hallquist & Dan Shallal"
date: "02 Sep 2026"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Inspecting and Diagnosing Pipeline Runs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

BrainGnomes separates routine progress inspection from failure diagnosis:

- `inspect_project()` answers, "What is happening now?"
- `diagnose_project()` answers, "What failed, and where are its logs?"

Both functions use the project's job-tracking database. Inspection does not
contact the scheduler, submit work, or change project files.

```{r, eval=FALSE}
library(BrainGnomes)
scfg <- load_project("/project/my_study")
```

# Inspect current project progress

The default inspection integrates all tracked runs:

```{r, eval=FALSE}
status <- inspect_project(scfg)
status
```

The printed dashboard gives overall counts, stage/stream progress, and a
bounded list of subjects that are active or need attention. Long tracking
fields such as scheduler options, paths, and manifests remain available in the
returned object but are not printed automatically.

The object contains ordinary data frames at several resolutions:

```{r, eval=FALSE}
status$overview
status$stages
status$subjects
status$subject_stages
status$runs
status$attempts
status$jobs
```

The same tables are available through `summary()`:

```{r, eval=FALSE}
summary(status)                         # one-row overview
summary(status, by = "stage")
summary(status, by = "subject")
summary(status, by = "subject_stage")
```

Because these are data frames, ordinary R queries work without another API:

```{r, eval=FALSE}
subset(status$subjects, status != "COMPLETED")
subset(status$subject_stages, sub_id == "540296")
subset(status$jobs, lifecycle_status %in% c("RUNNING", "QUEUED"))
```

## How status is integrated across runs

Each subject/session/stage/stream combination is treated as a logical work
unit. Related controller, array, and sentinel jobs are grouped into that unit.
For current project status, BrainGnomes uses the **newest attempt** for every
unit rather than the best status ever observed.

Consequently, a successful retry supersedes an older failure, while a failed
forced rerun is not hidden by an older success. Work omitted from a later
partial retry retains its newest earlier state. Historical attempts remain
available in `status$attempts` and `status$runs`.

Project-wide inspection describes work represented in the tracking database.
It does not invent a status for configured work that has never been submitted.

## Interpreting statuses

Raw database statuses are retained in `status$jobs$status`. Summarized views use
friendlier lifecycle labels:

- `COMPLETED`: all jobs in the work unit completed.
- `RUNNING`: at least one job has started and no failure takes precedence.
- `QUEUED`: submitted work has not started.
- `FAILED`: the work itself failed.
- `BLOCKED`: the database records `FAILED_BY_EXT`, meaning required upstream
  work failed.
- `CANCELLED`: the work was cancelled.
- `UNKNOWN`: the recorded state is missing or unrecognized.

Counts are always retained alongside the convenience status, so mixed states
remain visible.

# Choose the run you mean

Use the run table to find historical IDs:

```{r, eval=FALSE}
status$runs
run_id <- status$runs$run_id[[1L]]
```

For retry or cancellation, an explicit run ID is safer than `"latest"` because
another submission could become the newest run.

# Inspect one run without prompts

Supply `"latest"` or an explicit ID to restrict every inspection view to one
submission:

```{r, eval=FALSE}
latest <- inspect_project(scfg, run_id = "latest")
selected <- inspect_project(scfg, run_id = run_id)

selected$overview
selected$subjects
selected$subject_stages
selected$jobs
```

The older `get_project_runs()` and `get_run_jobs()` functions remain as
deprecated compatibility wrappers. New code should use `$runs` and `$jobs`
from `inspect_project()`.

# Diagnose failures

Routine monitoring belongs in `inspect_project()`. When current work has
failed, use `diagnose_project()` to collect the unresolved job records and
matching logs:

```{r, eval=FALSE}
diagnosis <- diagnose_project(scfg)
diagnosis$failures
diagnosis$logs
```

An existing inspection snapshot can be reused:

```{r, eval=FALSE}
diagnosis <- diagnose_project(status)
```

For a historical post-mortem, select the run explicitly:

```{r, eval=FALSE}
diagnosis <- diagnose_project(scfg, run_id = run_id)
```

The guided dependency and log browser is now reached through the same function:

```{r, eval=FALSE}
diagnose_project(scfg, run_id = run_id, interactive = TRUE)
```

`diagnose_pipeline()` remains temporarily as a deprecated compatibility wrapper
for interactive diagnosis.

# Inspect run provenance

Before deciding what to rerun, inspect the settings and scope that produced the
selected run:

```{r, eval=FALSE}
provenance <- get_run_provenance(scfg, run_id)
provenance$request
provenance$execution$subjects
provenance$configuration$snapshot_file
```

The provenance bundle records the resolved stages and subjects, configuration,
resources, software, host, and fingerprints of execution-driving files.

# Retry failed work after correcting the cause

A retry creates a **new run**. It does not restart scheduler jobs in place and
does not change the original run. Always preview it first:

```{r, eval=FALSE}
retry_plan <- retry_project_run(scfg, run_id, dry_run = TRUE)
retry_plan$jobs
retry_plan$subjects
```

After correcting the underlying cause, submit the retry explicitly:

```{r, eval=FALSE}
retry_run <- retry_project_run(scfg, run_id, dry_run = FALSE)
```

Jobs marked `FAILED_BY_EXT` were blocked by an earlier failure and are excluded
by default. Include them when appropriate:

```{r, eval=FALSE}
retry_plan <- retry_project_run(
  scfg, run_id, include_blocked = TRUE, dry_run = TRUE
)
```

# Cancel work that is still active

Cancellation applies only to tracked jobs that are queued or running. It does
not delete data, logs, or completed outputs. Preview the scheduler commands
before making a separate explicit cancellation request:

```{r, eval=FALSE}
cancel_project_run(scfg, run_id, dry_run = TRUE)
cancel_project_run(scfg, run_id, dry_run = FALSE)
```

# Command-line inspection and recovery

The command line exposes the same inspection resolutions:

```bash
BrainGnomes status /project/my_study
BrainGnomes status /project/my_study --view=stages
BrainGnomes status /project/my_study --view=subjects
BrainGnomes status /project/my_study --run=latest --watch
BrainGnomes status /project/my_study --run=<run-id> --view=jobs --format=csv

BrainGnomes diagnose /project/my_study
BrainGnomes diagnose /project/my_study --run=<run-id> --interactive
BrainGnomes logs /project/my_study --run=<run-id> --failed-only --tail=50
BrainGnomes provenance /project/my_study --run=<run-id> --format=json

BrainGnomes retry /project/my_study --run=<run-id> --dry-run
BrainGnomes retry /project/my_study --run=<run-id> --yes
BrainGnomes cancel /project/my_study --run=<run-id> --dry-run
BrainGnomes cancel /project/my_study --run=<run-id> --yes
```

Retry and cancellation require either a dry-run preview or explicit
confirmation. Neither routine inspection nor diagnosis changes scheduler state.
