---
title: "Local onboarding and prerequisites"
author: "Michael Hallquist"
date: "25 Aug 2026"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Local onboarding and prerequisites}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

# What works without a cluster?

BrainGnomes can be installed and loaded on an ordinary computer. Configuration
inspection, BIDS filename utilities, status-table handling, and native image
helpers do not submit jobs and do not require SLURM, TORQUE/PBS, or a container
runtime. Full project execution is different: `run_project()` coordinates HPC
jobs and therefore requires a supported scheduler, shared storage, and the
external software for every selected stage.

This walkthrough is evaluated when the vignette is built. It provides a first
successful task without assuming access to study data or a cluster.

```{r setup}
library(BrainGnomes)
```

# Inspect the installed example configuration

The package includes a miniature YAML configuration whose stages are disabled
and whose paths are explicit placeholders. Load it with `validate = FALSE` for
inspection only; validation is deliberately deferred because the placeholder
paths do not exist.

```{r inspect-config}
example_config_file <- system.file(
  "extdata", "example_project_config.yaml",
  package = "BrainGnomes"
)
stopifnot(nzchar(example_config_file))

example_config <- load_project(example_config_file, validate = FALSE)
c(
  project_name = example_config$metadata$project_name,
  scheduler = example_config$compute_environment$scheduler
)

stage_names <- c(
  "flywheel_sync", "bids_conversion", "mriqc", "fmriprep", "aroma",
  "postprocess", "extract_rois"
)
vapply(stage_names, function(stage) isTRUE(example_config[[stage]]$enable), logical(1))
```

Copy this file before adapting it:

```{r copy-config, eval=FALSE}
file.copy(example_config_file, "project_config.yaml")
scfg <- load_project("project_config.yaml", validate = FALSE)
scfg <- edit_project(scfg)
```

Do not submit the unchanged example: its `/path/to/...` values are not real.
Interactive `setup_project()` remains the recommended way to create a complete
configuration because it records only the stages and resources you choose.

# Exercise BIDS filename handling locally

These helpers operate on names and require neither image data nor external
software:

```{r bids-roundtrip}
filenames <- c(
  "sub-01_task-rest_space-MNI152NLin6Asym_desc-preproc_bold.nii.gz",
  "sub-02_ses-A_task-nback_run-2_bold.nii.gz"
)
bids_info <- extract_bids_info(filenames, drop_unused = TRUE)
bids_info
stopifnot(identical(construct_bids_filename(bids_info), filenames))
```

# Exercise a native image helper locally

`image_quantile()` is compiled with the package but does not invoke a scheduler
or container. Here it reads a tiny synthetic NIfTI image:

```{r local-image}
image_file <- tempfile(fileext = ".nii.gz")
RNifti::writeNifti(
  RNifti::asNifti(array(1:27, dim = c(3, 3, 3))),
  image_file
)
image_quantile(image_file, quantiles = c(0.25, 0.5, 0.75))
unlink(image_file)
```

# Understand dry runs

Once a real configuration exists, a dry run validates accessible paths and
selected stage resources, resolves stream settings, and reports planned
subject/session work without submitting jobs:

```{r dry-run, eval=FALSE}
scfg <- load_project("/real/project/project_config.yaml")
run_project(
  scfg,
  steps = c("postprocess", "extract_rois"),
  dry_run = TRUE
)
```

A dry run is submission-free, but it is not a substitute for configuration:
the selected stages still need real directories, inputs, containers, and other
required files so that validation can describe a trustworthy plan.

# Stage-specific prerequisites

| Stage | External runtime and inputs |
|---|---|
| Flywheel synchronization | Flywheel `fw` CLI and account access |
| DICOM-to-BIDS conversion | HeuDiConv container, DICOM inputs, and Python heuristic |
| BIDS validation | BIDS validator; configured with the project and submitted separately through `run_bids_validation()` |
| MRIQC | MRIQC container |
| fMRIPrep | fMRIPrep container, BIDS inputs, TemplateFlow cache, and FreeSurfer license |
| ICA-AROMA | fMRIPost-AROMA container |
| Postprocessing | FSL container; Python with `nibabel`, `nilearn`, and `templateflow` when template masks must be resampled |
| ROI extraction | Postprocessed BOLD inputs and compatible atlas/mask NIfTI files |

Every project-managed stage additionally requires SLURM or TORQUE/PBS, Bash,
shared writable storage, and site-appropriate scheduler settings. BrainGnomes
batch scripts invoke `singularity`; Apptainer is suitable when it supplies the
compatible `singularity` command.
