
```{r}
#| include: false
# Load the exported snapshot and prebuilt widgets. Rendering does not query
# project files, the tracking database, or a scheduler.
library(htmltools)
library(reactable)
library(plotly)
qc <- readRDS(params$snapshot)
widgets <- readRDS(params$widgets)
```

# Overview

## Row {height=140}

```{r}
#| content: valuebox
list(icon = "people", color = "#183044", title = "Participants in workflow scope",
     value = length(unique(na.omit(qc$workflow$sub_id))))
```

```{r}
#| content: valuebox
list(icon = "layers", color = "#127a70", title = "Available derivative records",
     value = paste(sum(qc$inventory$output_status == "available"), "/", nrow(qc$inventory)))
```

```{r}
#| content: valuebox
list(icon = "exclamation-circle", color = "#b84653", title = "Missing derivative records",
     value = sum(qc$inventory$output_status == "missing"))
```

```{r}
#| content: valuebox
list(icon = "activity", color = "#326eac", title = "Workflow units completed",
     value = paste(sum(qc$workflow$status == "COMPLETED"), "/", nrow(qc$workflow)))
```

## Row {height=440}

### Column {width=65%}

```{r}
#| title: Workflow progress
widgets$workflow_plot
```

### Column {width=35%}

```{r}
#| title: About this snapshot
tags$div(class = "qc-about",
  tags$p(class = "qc-eyebrow", "BRAINGNOMES / QC SNAPSHOT"),
  tags$h3(qc$metadata$project),
  tags$p(tags$strong("Collected: "), qc$metadata$created_at),
  tags$p(if (qc$metadata$scheduler_refreshed) "Active jobs compared with the scheduler at collection time."
    else "Database snapshot; the scheduler was not queried."),
  tags$p("Collect a new snapshot to refresh this dashboard."),
  tags$p(class = "qc-note", "Workflow completion, file availability, and computational QC checks answer different questions. None establishes that an image is scientifically acceptable."),
  tags$div(class = "qc-downloads",
    tags$a(href = "inventory.tsv", download = "inventory.tsv", "Derivative TSV"),
    tags$a(href = "workflow.tsv", download = "workflow.tsv", "Workflow TSV"),
    tags$a(href = "inventory.rds", download = "inventory.rds", "Complete R object"))
)
```

## Row {height=410}

::: {.card title="Workflow inventory"}

Recorded states are shown alongside configured work with no tracking record.
Units can cover a whole participant/session and several acquisitions. Missing
upstream inputs remain in this table before derivative paths can be resolved.

```{r}
widgets$workflow
```

:::

# Derivatives

## Row {height=620}

::: {.card title="Explore outputs and evidence"}

Search or filter any column, then expand a row for source files, diagnostics,
validation messages, and exact output-manifest associations. “Available” means
a nonempty file exists. “Attention” marks missing/empty outputs, failed or
invalid audits, recorded ROI loss, shared destinations, and failed workflow
context; it is not an exclusion recommendation.

```{r}
widgets$inventory
```

:::

## Row {height=350}

::: {.card title="Raw BOLD acquisitions"}

Raw files are listed separately from derivative records; echoes and acquisition
labels are retained. One acquisition can produce multiple spaces, streams, and
atlas outputs. Derivative counts therefore are not scan counts.

```{r}
widgets$acquisitions
```

:::

# QC metrics

## Row {height=530}

::: {.card title="Motion and retained data"}

Choose a metric/stage/stream/task/atlas group from the plot menu. Hover for
values, units, and observed/missing denominators. Each evidence source appears
once within its group. Values are descriptive; no dashboard cutoff is applied.

```{r}
widgets$metric_plot
```

:::

## Row {height=400}

::: {.card title="Metric definitions and sources"}

FD comes from the original fMRIPrep confounds. Censor retention describes the
saved mask and does not establish whether volumes were removed. ROI retention
uses the extraction rule saved in the diagnostics. Missing diagnostics remain
unavailable, rather than being interpreted as a passing result.
MRIQC metrics retain their upstream names and definitions in a separate group.

```{r}
widgets$metrics
```

:::

## Row {height=320}

::: {.card title="Postprocessing validation checks"}

Audits test computational operations. Their timestamps and source files are
retained in the evidence table; an older audit may outlive its derivative.

```{r}
widgets$checks
```

:::

## Row {height=350}

::: {.card title="Regional voxel coverage"}

Filter by participant, stream, atlas, or retained status to locate individual
ROIs with limited usable voxels. The saved reason explains whether a label
fell outside the mask, contained unusable BOLD signal, or missed its threshold.

```{r}
widgets$regions
```

:::

# Workflow & evidence

## Row {.tabset height=520}

::: {.card title="Job attempts"}

Expand a job for stdout/stderr, its submission manifest, and runtime receipt.
Historical attempts are retained. A newer attempt does not establish the origin
of an existing derivative.

```{r}
widgets$jobs
```

:::

::: {.card title="Active jobs"}

```{r}
widgets$active
```

:::

::: {.card title="Scheduler comparison"}

```{r}
widgets$reconciliation
```

:::

::: {.card title="Output manifest checks"}

These checks compare exact paths, sizes, and modification times against saved
job manifests. They do not verify file contents or scientific quality.

```{r}
widgets$manifests
```

:::

## Row {.tabset height=440}

::: {.card title="Evidence library"}

Links open original reports and files. Keep the source project accessible;
moving this dashboard alone does not copy upstream evidence or its assets.

```{r}
widgets$evidence
```

:::

::: {.card title="Collection issues"}

Malformed diagnostics and unresolved associations appear here. Other available
evidence remains accessible; a collection problem does not imply a QC pass.

```{r}
widgets$issues
```

:::

::: {.card title="Downloads & interpretation"}

- [Raw acquisitions](acquisitions.tsv)
- [Derivative inventory](inventory.tsv)
- [Workflow inventory](workflow.tsv)
- [Metric values, definitions, and sources](metrics.tsv)
- [Regional voxel coverage](regions.tsv)
- [Validation checks](checks.tsv)
- [Evidence paths and timestamps](evidence.tsv)
- [Job history](jobs.tsv) and [logical attempts](attempts.tsv)
- [Output manifest checks](manifest_files.tsv)
- [Collection issues](issues.tsv)
- [Complete R object](inventory.rds) and [snapshot metadata](snapshot.json)

All downloads describe the same snapshot. Expected native outputs are resolved
from the **current configuration and discovered upstream inputs**. Until inputs
exist, workflow expectations provide the denominator; output spaces and echo
combinations are not guessed. Configurations may have changed since a historical
attempt. Consult its saved manifest and receipt for its execution settings.

This dashboard supports inspection and discussion. It does not collect human
ratings, make inclusion decisions, or modify the pipeline.

:::
